# Reading in Labelled Data: Stata {#chp-2}

The first step to working with labeled data in R is loading a labeled data file, and making sure the labels are attached. Several of the packages in [Chapter 1](#chp-1) have functions for loading labeled data while others have functions checking for labels. A few packages do both. The first table shows which packages read .dta files into R and their functions that do so. The second table shows which packages allow you to checkout the labels of the loaded data.

+---------------+--------------------------------+
| Package       | Function for loading .dta file |
+===============+================================+
| `haven`       | `read_dta()`                   |
+---------------+--------------------------------+
| `sjlabelled`  | `read_stata()`                 |
+---------------+--------------------------------+
| `foreign`     | `read.dta()` Up to 12          |
+---------------+--------------------------------+
| `readstata13` | `read.dta13()` Past 12         |
+---------------+--------------------------------+
Table: (\#tab:read-dta) Functions to read in Stata data


+---------------+------------------+---------------------+-------------------+
| Package       | Labels Exist     | View Variable Label | View Value Labels |
+===============+==================+=====================+===================+
| `labelled`    | `is.labelled()`  | `var_label()`       | `val_labels()`    |
+---------------+------------------+---------------------+-------------------+
| `sjlabelled`  | `is_labelled()`  | `get_label()`       | `get_labels()`    |
+---------------+------------------+---------------------+-------------------+
| `expss`       | `is.labelled()`  | `var_lab()`         | `val_lab()`       |
+---------------+------------------+---------------------+-------------------+
| `readstata13` |  N/A             | `varlabel()`        | `get.label()`     |
+---------------+------------------+---------------------+-------------------+
Table: (\#tab:labels-table) Functions to look at labels

Stata can attach labels both to the variable name and the variable values. Packages that work with labels tend to check for variable and value labels separately.

This chapter is organized as follows. Each section reads in .dta files according to a package: `haven`, `sjlabelled`, and `readstata13`. The `foreign` package is not covered as `foreign::read.dta()` is frozen and will not be updated to support Stata formats after 12. Once the data is loaded, we look at the class of the data as well as the class and structure of a variable that *should* be labeled (hint: not all packages retain labels of the data). The data used is City Temps, an example data frame found in Stata. The variable `division` has labeled numeric values. To follow along, you can find the structure of the data in the [Appendix](#appendix).

Each section has a subsection that fills in the table of functions to look at labels. That is, a variable that should be labeled is passed into the function in the table and TRUE, FALSE, or ERROR is recorded in the table as a result. A FALSE result means labels should appear, but the function returns NULL. A ERROR result means even if the function is given proper syntax, an error occurs. This happens mostly commonly when the data is not of an appropriate class for the function.

A note: The `haven` and `labelled` packages were written to work together, and thus inherit some functions from each other. In particular, `labelled::is.labelled()` is identical to `haven::is.labelled()`. Therefore, even though `haven` is a separate package, when it comes to checking for the existence of labels, we will only use `labelled` to avoid redundancy.

Another note: If you do not want to gain a deeper understanding of how the various packages treat labeled data and just want to see which functions work with which data inputs, you can use the following links to skip to the answers.

* [Answers to Tabel 2 with data read in with `haven::read_dta()`](\#tab:haven-see-labels)

## Loading .dta file with `haven`

The CityTemp_haven data structure can be copied from [here](#appendix-CT-haven).

```{r haven-loading-dta}

CityTemp_haven <- haven::read_dta("CityTemp.dta")

names(attributes(CityTemp_haven))
class(CityTemp_haven)

attributes(CityTemp_haven$division)
str(CityTemp_haven$division)


```

Since the `haven` package is part of the tidyverse, `read_dta()` naturally reads in the data as a tibble. The variable `division` has four attributes: label, format.stata, class, and labels. Importantly, the variable belongs to the `"haven_labelled"` class. The `"haven_labelled"` class is another product of the tidyverse, so functions from outside the tidyverse may not work very well.

### Checking labels of `haven` loaded data

#### Label Existence {#label-exist-haven}
```{r haven-loaded-labels-exist}

labelled::is.labelled(CityTemp_haven$division)
sjlabelled::is_labelled(CityTemp_haven$division)
expss::is.labelled(CityTemp_haven$division)
```

According to the packages `labelled` and `sjlabelled`, the `division` variable is labeled, but the `expss` package cannot find any. Why? Let's look at what these functions are looking for.

```{r label-existence}

labelled::is.labelled

sjlabelled::is_labelled

expss::is.labelled

```

The function `labelled::is.labelled()` returns TRUE if the object is of the class `"haven_labelled"`. The function `sjlabelled::is_labelled()` returns TRUE if the object is of class `"haven_labelled"` **or** `"labelled"`. The function `expss::is.labelled` returns TRUE if the object is of the `"labelled"`. When we read in the data, we saw that the variable `division` belongs to the class `"haven_labelled"`, which causes `labelled::is.labelled()` and `sjlabelled::is_labelled()` to return TRUE, and `expss::is.labelled()` to return FALSE.

The next step is to try to see the variable and value labels.

#### Variable Labels {#var-lab-haven}

```{r haven-loaded-var-labels,error=TRUE}

labelled::var_label(CityTemp_haven$division)
sjlabelled::get_label(CityTemp_haven$division)
expss::var_lab(CityTemp_haven$division)
readstata13::varlabel(CityTemp_haven,"division")
```

Both the `labelled` and `sjlabelled` packages told us the variable `division` is labeled. It makes sense then that we can use their functions to look at the variable's labels. The `expss` function, on the other hand, told us the variable was not labeled, and yet we can still see the labels.

It takes a little investigating, but eventually we can find that `labelled::var_label()`, `sjlabelled::get_label()`, and `expss::var_lab()` all fundamentally call the same method: `attr(x, "label", exact = TRUE)`. As long as an object's label is stored in an attribute called `"label"`, any of these functions will return the label.

The functions in the `readstata13` package are formulated differently. Trying to look at the variable label using `readstata13::varlabel()` throws an error. We can take a closer look at the function to see why.

```{r readstata13-varlabel-haven}

readstata13::varlabel

```

The third lines of the function is `varlabel <- attr(dat, "var.labels")`. We know from looking at `division`'s attributes earlier that it does not have an attribute called `"var.labels"`. When `attr(CityTemp, "var.labels")` is evaluated, the result is `r attr(CityTemp_haven, "var.labels")`, hence the error message `attempt to set an attribute on NULL`.


#### Value Labels {#val-labl-haven}

```{r haven-loaded-value-labels,error=TRUE}

labelled::val_labels(CityTemp_haven$division)
sjlabelled::get_labels(CityTemp_haven$division)
expss::val_lab(CityTemp_haven$division)
readstata13::get.label(CityTemp_haven,"division")
```

The same pattern seen for variable labels is repeated for value labels. The functions from the packages `labelled`, `sjlabelled`, and `expss` all return the value labels, even though `expss::is.labelled()` returned FALSE. While `readstata13::get.label()` does not throw an error, it returns `NULL`.

The functions `labelled::val_labels()`, `sjlabelled::get_labels()`, and `expss::val_lab()` are internally very similair to their variable label function counterparts. All three functions call `attr(x, "labels", extract = TRUE)`. The function `sjlabelled::get_labels()` further processes the attributes, resulting in a slightly different output than `labelled::val_labels()` and `expss::val_lab()`.  Notably, `labelled::val_labels()` only does this if the object belongs to either the `"haven_labelled"` class or the `"data.frame"` class. Otherwise, it will just return `NULL`.


The reason behind `readstata13::get.label()` returning `NULL` is similar to its variable label counter part. Let's look at the function.


```{r readstata13-get.label-haven}

readstata13::get.label

```

It turns out `readstata13::get.label()` is a straightforward wrapper function to call the label of a specific variable from the data attribute list. Again, we know that the variable `division` from the `CityTemp_haven` data does not have an attribute named `"label.table"`. When we run `attr(CityTemp_haven, "label.table")`, we get `r attr(CityTemp_haven, "label.table")`. Any indexing of `NULL` will simply return `NULL`. 


With that, we can fill in our table for data read into R using the `haven` package.


+---------------+------------------+---------------------+-------------------+
| Package       |  Labels Exist    | View Variable Label | View Value Labels |
+===============+=+================+=====================+===================+
| `labelled`    | TRUE             | TRUE                | TRUE              |
+---------------+------------------+---------------------+-------------------+
| `sjlabelled`  | TRUE             | TRUE                | TRUE              |
+---------------+------------------+---------------------+-------------------+
| `expss`       | FALSE            | TRUE                | TRUE              |
+---------------+------------------+---------------------+-------------------+
| `readstata13` |  N/A             | ERROR               | FALSE             |
+---------------+------------------+---------------------+-------------------+
Table: (\#tab:haven-see-labels) [Table 2](\#tab:labels-table) filled in with data read in from `haven::read_dta()`





## Loading .dta file with `sjlabelled`

The `CityTemp_sjlabelled` data structure can be copied from [here](#appendix-CT-sjlabelled).

```{r sjlabelled-loading-dta}

CityTemp_sjlabelled <- sjlabelled::read_stata("CityTemp.dta")

names(attributes(CityTemp_sjlabelled))
class(CityTemp_sjlabelled)

attributes(CityTemp_sjlabelled$division)
str(CityTemp_sjlabelled$division)

```

The `sjlabelled` package is not part of the tidyverse. We see the data read in by `sjlabelled::read_stata()` only belongs to the class `"data.frame"`. It is not a tibble. The variable `division` again as four attributes. The attributes are similar to those of the variable `division` when the data set is read in with `haven::read_dta()`. There are two notable differences. First, the class is `"factor"`, whereas the `division` variable from the `CityTemp_haven` belonged to the classes `"haven_labelled"`, `"vctrs_vctr"`, and `"double"`. Next, since the variable is a factor, it has an attribute `levels`. The `levels` attribute of `division` from `CityTemp_sjlabelled` takes the place of `format.stata` attribute that `division` from `CityTemp_haven` has. An important similarity between the data set variables is that they both have two attributes named `label` and `labels`. With this knowledge, a reader could reasonably fill out their expectations for [Table 2](\#tab:labels-table) with `sjlabelled` loaded data.


### Checking labels of `sjlabelled` loaded data

#### Label Existence {#label-exist-sjlabelled}

```{r sjlabelled-loaded-labels-exist}

labelled::is.labelled(CityTemp_sjlabelled$division)
sjlabelled::is_labelled(CityTemp_sjlabelled$division)
expss::is.labelled(CityTemp_sjlabelled$division)

```

All three functions return FALSE. This should not be surprising, since the variable `division` in this case does not belong to either the `"haven_labelled"` class or the `"labelled"` class. Thus, even when the user the `sjlabelled` package to load the data and check for labels, `sjlabelled` will return that the data is not labeled.

#### Variable Labels {#var-lab-sjlabelled}
```{r sjlabelled-loaded-var-labels,error=TRUE}
## Variable Labels
labelled::var_label(CityTemp_sjlabelled$division)
sjlabelled::get_label(CityTemp_sjlabelled$division)
expss::var_lab(CityTemp_sjlabelled$division)
readstata13::varlabel(CityTemp_sjlabelled,"division")
```

Just as with the `haven` loaded data, the first three functions return the variable label, while the `readstata13::varlabel()` throws an error. The reasoning is exactly the same as explained in section (#var-lab-haven): `labelled::var_label()`, `sjlabelled::get_label()`, and `expss::var_lab()` all look for an attribute `label`, while `readstata13::varlabel()` looks for an attribute `var.label`. The former exists while the latter does not.

#### Value Labels {#val-lab-sjlabelled}

```{r sjlabelled-loaded-value-labels,error=TRUE}
## Value Labels
labelled::val_labels(CityTemp_sjlabelled$division)
sjlabelled::get_labels(CityTemp_sjlabelled$division)
expss::val_lab(CityTemp_sjlabelled$division)
readstata13::get.label(CityTemp_sjlabelled,"division")

```

In a turn of events, `sjlabelled::get_labels()` and `expss::val_lab()` return value labels, while both `labelled::val_labels()` and `readstata13::get.label()` return `NULL`. But... the functions from `sjlabelled`, `expss` *and* `labelled` are supposed to return the value of the `labels` attribute, right? WRONG. Recall: `labelled::val_labels()` will only return the value of the object's `labels` attribute **if the object belongs to the `"haven_labelled"` or `"data.frame"` class.** Otherwise, it will return `NULL`, as it does in this case. The variable `division` from the `CityTemp_sjlabelled` data only belongs to the `"factor"` class, so `labelled::val_labels()` returns `NULL`.

The function `readstata13::get.label()` runs into the same problem as before; namely, the data `CityTemp_sjlabelled` does not have an attribute `"label.table"`.

Now, we can fill in [Table 2](\#tab:labels-table) for data read in by `sjlabelled::read_stata()`.

+---------------+------------------+---------------------+-------------------+
| Package       |  Labels Exist    | View Variable Label | View Value Labels |
+===============+=+================+=====================+===================+
| `labelled`    | FALSE            | TRUE                | FALSE             |
+---------------+------------------+---------------------+-------------------+
| `sjlabelled`  | FALSE            | TRUE                | TRUE              |
+---------------+------------------+---------------------+-------------------+
| `expss`       | FALSE            | TRUE                | TRUE              |
+---------------+------------------+---------------------+-------------------+
| `readstata13` |  N/A             | ERROR               | FALSE             |
+---------------+------------------+---------------------+-------------------+




## Loading .dta file with `readstata13`


```{r readstata13-loading-CT}

CityTemp_readstata13 <- readstata13::read.dta13("CityTemp.dta")

names(attributes(CityTemp_readstata13))
class(CityTemp_readstata13)

attributes(CityTemp_readstata13$division)
str(CityTemp_readstata13$division)

```


### Checking labels of `readstata13` loaded data

```{r readstata13-loaded-label-checks-CT}
# Labels Exist?
labelled::is.labelled(CityTemp_readstata13$division)
sjlabelled::is_labelled(CityTemp_readstata13$division)
expss::is.labelled(CityTemp_readstata13$division)

# View Labels
labelled::var_label(CityTemp_readstata13$division)
labelled::val_labels(CityTemp_readstata13$division)

sjlabelled::get_label(CityTemp_readstata13$division)
sjlabelled::get_labels(CityTemp_readstata13$division)

expss::var_lab(CityTemp_readstata13$division)
expss::val_lab(CityTemp_readstata13$division)

readstata13::get.label.name(CityTemp_readstata13)
readstata13::get.label(CityTemp_readstata13)


```


+---------------+--------------+---------------------+-------------------+
| Package       | Labels Exist | View Variable Label | View Value Labels |
+===============+=+============+=====================+===================+
| `labelled`    | FALSE        | FALSE               | FALSE             |
+---------------+--------------+---------------------+-------------------+
| `sjlabelled`  | FALSE        | FALSE               | TRUE              |
+---------------+--------------+---------------------+-------------------+
| `expss`       | FALSE        | FALSE               | FALSE             |
+---------------+--------------+---------------------+-------------------+
| `readstata13` |  N/A         | TRUE                | TRUE              |
+---------------+--------------+---------------------+-------------------+


## Function Cohesion Summary
### Variable Labels
+-----------------------------+-------------------------+---------------------------+--------------------+-------------------------+
| City Temp Data              | `labelled::var_label()` | `sjlabelled::get_label()` | `expss::var_lab()` | `readstata13::varlab()` |
+=============================+=========================+===========================+====================+=========================+
| `haven::read_dta()`         | TRUE                    | TRUE                      | TRUE               | ERROR                   |
+-----------------------------+-------------------------+---------------------------+--------------------+-------------------------+
| `sjlabelled::read_stata()`  | TRUE                    | TRUE                      | TRUE               | ERROR                   |
+-----------------------------+-------------------------+---------------------------+--------------------+-------------------------+
| `readstata13::read.dta13()` | FALSE                   | FALSE                     | FALSE              | FALSE                   |
+-----------------------------+-------------------------+---------------------------+--------------------+-------------------------+

### Value Labels
+---------------------------- +--------------------------+----------------------------+--------------------+----------------------------+
| City Temp Data              | `labelled::val_labels()` | `sjlabelled::get_labels()` | `expss::val_lab()` | `readstata13::get.label()` |
+=============================+==========================+============================+====================+============================+
| `haven::read_dta()`         | FALSE                    | TRUE                       | TRUE               | FALSE                      |
+-----------------------------+--------------------------+----------------------------+--------------------+----------------------------+
| `sjlabelled::read_stata()`  | FALSE                    | TRUE                       | TRUE               | FALSE                      |
+-----------------------------+--------------------------+----------------------------+--------------------+----------------------------+
| `readstata13::read.dta13()` | FALSE                    | FALSE                      | FALSE              | FALSE                      |
+-----------------------------+--------------------------+----------------------------+--------------------+----------------------------+

