# Reading in Labelled Data: SPSS


+---------------+------------+-------------------+-------------+
| Package       | Loads .sav | Checks for Labels | View Labels |
+===============+============+===================+=============+
| `haven`       |  Yes       | Yes$^*$           | No          |
+---------------+------------+-------------------+-------------+
| `labelled`    | No         | Yes$^*$           | Yes         |
+---------------+------------+-------------------+-------------+
| `sjlabelled`  |  Yes       | Yes               | Yes         |
+---------------+------------+-------------------+-------------+
| `expss`       | Yes        | Yes               | Yes         |
+---------------+------------+-------------------+-------------+
| `foreign`     | Yes        | No                | No          |
+---------------+------------+-------------------+-------------+
| `readstata13` | No         | No                | Yes         |
+---------------+------------+-------------------+-------------+
| `Hmisc`       | No         | No                | Yes         |
+---------------+------------+-------------------+-------------+


$^*$The function used to check for labels in the `haven` and `labelled` packages is the same.



```{r drive-sav-path,echo=FALSE}

path_to_sav <- "/Volumes/GoogleDrive/Shared drives/Arab Barometer/AB7/Data/RELEASE_DATA/Latest/AB7_ENG_Release_Version6.sav"

```


## Loading .sav files

### Loading .sav file with haven

```{r haven-loading-sav}

public_data_sav <- haven::read_sav(path_to_sav)

class(public_data_sav)
class(public_data_sav$Q101)

```

### Loading .sav file with `sjlabelled`

```{r sjlabelled-loading-sav}

AB7_sjlabelled_sav <- sjlabelled::read_spss(path_to_sav)

class(AB7_sjlabelled_sav)
class(AB7_sjlabelled_sav$Q101)

```

### Loading .sav file with `expss`

```{r expss-loading-sav}

AB7_expss_sav <- expss::read_spss(path_to_sav)

class(AB7_expss_sav)
class(AB7_expss_sav$Q101)

```
